Automating Free Logic in HOL, with an Experimental Application in Category Theory

### 2018-12-17 FreeLogic/2018-JAR/article.tex
update

\author{Christoph Benzm\"uller\thanks{Benzmüller received funding from the German National Research
  Foundation DFG under Heisenberg grant \emph{Towards Computational
    Metaphysics} (BE 2501/9-2) and from VolkswagenStiftung under grant
  Consistent Rational Argumentation in Politics (CRAP).}  and Dana S. Scott}
\institute{Christoph Benzm\"uller \at
   Freie Universit\"at Berlin, Berlin, Germany \& University of Luxembourg, Luxembourg\\
%   Stanford University, CSLI/Cordura Hall, Stanford, CA 94305-4115, U.S. \\
   \email{c.benzmueller@gmail.com} \\
% \\
 %            \emph{Present address:} of F. Author  %  if needed
Dana S. Scott \at
Visiting Scholar at University of California, Berkeley,  USA \\
  \email{dana.scott@cs.cmu.edu}
}

@@

\begin{defn}[Syntax of $\HOL$]\label{holgrammar}
The terms of $\HOL$ with primitive equality and
definite description are defined by the following
grammar:\footnote{It is well known that we could work with a much smaller set
  of logical connectives, see e.g. \S1.4 of Andrews's overview
  article~\cite{AndrewsSEP}. The choice here closely reflects the
  set of primitive connectives as chosen in higher-order automated theorem provers such
  as LEO-II \cite{LEO}, Leo-III \cite{Leo-III}, and Satallax \cite{Satallax}.}
\begin{align*} 
  \hol{s},\hol{t} \quad ::= \quad & \hol{p_\alpha} \mid \hol{X_\alpha} \mid \hol{(\lam{X_\alpha}
  s_\beta)_{\alpha\ar\beta}} \mid \hol{(s_{\alpha\ar\beta}\,
                                    t_\alpha)_\beta} \mid
                                    \hol{s_\alpha = t_\alpha} \mid
                                    \hol{\neg_{o\ar o}\, s_o} \mid\\
  & \hol{(({\vee_{o\ar o\ar o}}\,s_o)\,t_o)} \mid \hol{\forall_{(\alpha\ar
    o)\ar o}(\lam{X_\alpha} s_o)} \mid \hol{\Descr(\lam{X_\alpha} s_o)}
\end{align*}
where $\hol{\alpha},\hol{\beta}\in \hol{T}$. $\hol{p_\alpha}$ denotes typed constants and
$\hol{X_\alpha}$ typed variables (distinct from $\hol{p_\alpha}$).  Complex typed
terms are constructed via abstraction and application. The type of
each term is given as a subscript.  Terms $\hol{s_o}$ of type $\hol{o}$ are called
\emph{formulas}.  The \emph{logical connectives} of choice are
$\hol{\neg_{o\ar o}}$, $\hol{\vee_{o\ar
  o\ar o}}$, $\hol{=_{\alpha\ar\alpha\ar  o}}$, $\hol{\forall_{(\alpha\ar
  o)\ar o}}$ and $\hol{\Descr_{(\alpha\ar
  o)\ar \alpha}}$ (where $\hol{\alpha\in T}$).  Type subscripts may be dropped if
irrelevant or obvious. Similarly, parentheses may be avoided.  Binder
notation $\hol{\all{X_\alpha} s_o}$ and $\hol{\Descr {X_\alpha}\lambdot s_o}$ is used as shorthand for
$\hol{\forall_{(\alpha\ar o)\ar o}
(\lam{X_\alpha} s_o)}$ and $\hol{\Descr(\lam{X_\alpha} s_o)}$, and infix notation $\hol{s \vee t}$ is employed
instead of $\hol{((\vee s)\, t)}$. From the above connectives, other logical
connectives, such as $\hol{\top}$, $\hol{\bot}$, $\hol{\wedge}$,
$\hol{\Impl}$, $\hol{\equiv}$ and $\hol{\exists}$, can be defined in
the usual way. For example, $\hol{\Impl}$ can be defined as $\hol{\lam{X_o}\lam{Y_o}
  \neg X \vee Y}$ and $\hol{\wedge}$ as $\hol{\lam{X_o}\lam{Y_o}
  \neg (\neg X \vee \neg Y)}$.
\end{defn}

@@

We can prove that the $i$ in axiom $C_i$ and the $j$ in axiom $D_i$
are unique. The proofs and the dependencies can be 
   found automatically by Sledgehammer.
$$  \forall y\lambdot \exists i\lambdot \id i \wedge i \comp y \cong y
\wedge (\forall j\lambdot d(\id j \wedge j \comp y \cong y) \imp i \cong j)
\quad 
    (\text{by } A_i, C_i, S_i) $$
$$ \forall x\lambdot \exists j\lambdot \id j \wedge x \comp j \cong x \wedge
(\forall i\lambdot (\id i \wedge x \comp i \cong x) \imp j \cong i) \quad   
    (\text{by }  A_i, D_i, S_i) $$

@@

As we will demonstrate now, the consequences of this version of the axiom are
even stronger. It implies that \emph{all} objects are defined,
that is, composition (as well as $dom$ and $cod$) become total operations.
The theory described by these axioms ``collapses'' to the theory of
monoids: If all objects are defined, then one can conclude from $A1$ that 
$\dom x \cong \dom y$ (resp.~$\dom x \cong \cod y$) and $\cod x \cong \cod y$), 
and according to 1.14 of \cite{FreydScedrov90}, 
the category reduces to a monoid provided that it is not empty.

@@

 Let $a\in D$ be an undefined object, that is, assume $\neg E a$.  By
 instantiating axiom $A3a$ with $a$ we have $a \comp (\dom a) \cong a$.  From
 this and definition of $\cong$ we know that $a \comp (\dom a)$ is not
 defined. This is easy to see, since if $a \comp (\dom a)$ were defined, we
 also had that $a$ is defined, which is not the case by assumption.
  Hence, $\neg E (a \comp (\dom a)$.
Next, we instantiate $A1$ with $a$ and $\dom a$ to obtain
   $E(a \comp (\dom a)) \biimp \dom a \cong \cod(\dom a)$. Moreover,
   by instantiating $A2a$ with $a$ we obtain $\cod(\dom a) \cong \dom a$,
   which we use (modulo symmetry and transitivity of $\cong$) to
   rewrite the former result into 
   $E(a \comp (\dom a)) \biimp \dom a \cong \dom a$. By reflexivity of
   $\cong$ we thus get $E(a \comp (\dom a))$, i.e. that $a \comp
   (\dom a)$ is defined, which contradicts $\neg E (a \comp (\dom
   a)$. $\Box$

@@

\bibitem{makarenko16:_autom_logik_logik_stufe}
Makarenko, I.: Automatisierung von freier logik in logik h{\"o}herer stufe
  (2016).
\newblock Bachelorarbeit, Institut f{\"u}r Informatik, Freie Universit{\"a}t
  Berlin

>>>>

\author{Christoph Benzm\"uller\thanks{Benzmüller received funding from the German National Research
  Foundation DFG under Heisenberg grant \emph{Towards Computational
    Metaphysics} (BE 2501/9-2) and from VolkswagenStiftung under grant
  \emph{Consistent Rational Argumentation in Politics} (CRAP).}  and Dana S. Scott}
\institute{Christoph Benzm\"uller \at
   Freie Universit\"at Berlin, Berlin, Germany \& University of Luxembourg, Luxembourg\\
%   Stanford University, CSLI/Cordura Hall, Stanford, CA 94305-4115, U.S. \\
   \email{c.benzmueller@gmail.com} \\
% \\
 %            \emph{Present address:} of F. Author  %  if needed
Dana S. Scott \at
Visiting Scholar at University of California, Berkeley,  USA \\
  \email{dana.scott@cs.cmu.edu}
}

@@

\begin{defn}[Syntax of $\HOL$]\label{holgrammar}
The terms of $\HOL$ with primitive equality and
definite description are defined by the following
grammar:\footnote{It is well known that we could work with a much smaller set
  of logical connectives, see e.g. \S1.4 of Andrews's overview
  article~\cite{AndrewsSEP}. The choice here closely reflects the
  set of primitive connectives as chosen in higher-order automated theorem provers such
  as LEO-II \cite{LEO}, Leo-III \cite{Leo-III}, and Satallax \cite{Satallax}.}
\begin{align*} 
  \hol{s},\hol{t} \quad ::= \quad & \hol{p_\alpha} \mid \hol{X_\alpha} \mid \hol{(\lam{X_\alpha}
  s_\beta)_{\alpha\ar\beta}} \mid \hol{(s_{\alpha\ar\beta}\,
                                    t_\alpha)_\beta} \mid
                                    \hol{s_\alpha = t_\alpha} \mid
                                    \hol{\neg_{o\ar o}\, s_o} \mid\\
  & \hol{(({\vee_{o\ar o\ar o}}\,s_o)\,t_o)} \mid \hol{\forall_{(\alpha\ar
    o)\ar o}(\lam{X_\alpha} s_o)} \mid \hol{\Descr_{(\alpha\ar
  o)\ar \alpha}(\lam{X_\alpha} s_o)}
\end{align*}
where $\hol{\alpha},\hol{\beta}\in \hol{T}$. $\hol{p_\alpha}$ denotes typed constants and
$\hol{X_\alpha}$ typed variables (distinct from $\hol{p_\alpha}$).  Complex typed
terms are constructed via abstraction and application. The type of
each term is given as a subscript.  Terms $\hol{s_o}$ of type $\hol{o}$ are called
\emph{formulas}.  The \emph{logical connectives} of choice are
$\hol{\neg_{o\ar o}}$, $\hol{\vee_{o\ar
  o\ar o}}$, $\hol{=_{\alpha\ar\alpha\ar  o}}$, $\hol{\forall_{(\alpha\ar
  o)\ar o}}$ and $\hol{\Descr_{(\alpha\ar
  o)\ar \alpha}}$ (where $\hol{\alpha\in T}$).  Type subscripts may be dropped if
irrelevant or obvious. Similarly, parentheses may be avoided.  Binder
notation $\hol{\all{X_\alpha} s_o}$ and $\hol{\Descr {X_\alpha}\lambdot s_o}$ is used as shorthand for
$\hol{\forall 
(\lam{X_\alpha} s_o)}$ and $\hol{\Descr(\lam{X_\alpha} s_o)}$, and infix notation $\hol{s \vee t}$ is employed
instead of $\hol{((\vee s)\, t)}$. From the above connectives, other logical
connectives, such as $\hol{\top}$, $\hol{\bot}$, $\hol{\wedge}$,
$\hol{\Impl}$, $\hol{\equiv}$ and $\hol{\exists}$, can be defined in
the usual way. For example, $\hol{\Impl}$ can be defined as $\hol{\lam{X_o}\lam{Y_o}
  \neg X \vee Y}$ and $\hol{\wedge}$ as $\hol{\lam{X_o}\lam{Y_o}
  \neg (\neg X \vee \neg Y)}$.
\end{defn}

@@

We can prove that the $i$ in axiom $C_i$ and the $j$ in axiom $D_i$
are unique. The proofs and the dependencies can be 
   found automatically by Sledgehammer.
$$  \forall y\lambdot \exists i\lambdot \id i \wedge i \comp y \cong y
\wedge (\forall j\lambdot (\id j \wedge j \comp y \cong y) \imp i \cong j)
\quad 
    (\text{by } A_i, C_i, S_i) $$
$$ \forall x\lambdot \exists j\lambdot \id j \wedge x \comp j \cong x \wedge
(\forall i\lambdot (\id i \wedge x \comp i \cong x) \imp j \cong i) \quad   
    (\text{by }  A_i, D_i, S_i) $$

@@

As we will demonstrate now, the consequences of this version of the axiom are
even stronger. It implies that \emph{all} objects are defined,
that is, composition (as well as $dom$ and $cod$) become total operations.
The theory described by these axioms ``collapses'' to the theory of
monoids: If all objects are defined, then one can conclude from $A1$ that 
$\dom x \cong \dom y$ (resp.~$\dom x \cong \cod y$ and $\cod x \cong \cod y$), 
and according to 1.14 of \cite{FreydScedrov90}, 
the category reduces to a monoid provided that it is not empty.

@@

 Let $a\in D$ be an undefined object, that is, assume $\neg E a$.  By
 instantiating axiom $A3a$ with $a$ we have $a \comp (\dom a) \cong a$.  From
 this and definition of $\cong$ we know that $a \comp (\dom a)$ is not
 defined. This is easy to see, since if $a \comp (\dom a)$ were defined, we
 also had that $a$ is defined, which is not the case by assumption.
  Hence, $\neg E (a \comp (\dom a))$.
Next, we instantiate $A1$ with $a$ and $\dom a$ to obtain
   $E(a \comp (\dom a)) \biimp \dom a \cong \cod(\dom a)$. Moreover,
   by instantiating $A2a$ with $a$ we obtain $\cod(\dom a) \cong \dom a$,
   which we use (modulo symmetry and transitivity of $\cong$) to
   rewrite the former result into 
   $E(a \comp (\dom a)) \biimp \dom a \cong \dom a$. By reflexivity of
   $\cong$ we thus get $E(a \comp (\dom a))$, i.e. that $a \comp
   (\dom a)$ is defined, which contradicts $\neg E (a \comp (\dom
   a))$. $\Box$

@@

\bibitem{makarenko16:_autom_logik_logik_stufe}
Makarenko, I.: Automatisierung von Freier Logik in Logik h{\"o}herer Stufe
  (2016).
\newblock Bachelorarbeit, Institut f{\"u}r Informatik, Freie Universit{\"a}t
  Berlin
